Estimating duration in the presence of misclassification

### 2024-02-05 biology-data.tex
proof read

In this thesis, I make use of data from two studies which recruited participants from the general population in the UK.
Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
Therefore, the results from these studies are more generalisable to the entire UK population.
The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) performed daily testing of a small number of individuals with limited follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest, but the results are limited because the sparse sampling means that the viral load dynamics cannot be estimated.
In particular, the sparse sampling means that the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.

@@

@@

The \emph{positivity rate} is the proportion of people in a given population that (if they were tested) would test positive for SARS-CoV-2 at a defined point in time; in this context test means an RT-PCR test~\autocite{cisMethodsONS}.
A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
Larger time periods (\ie weeks instead of days) allow more granular estimates on other variables, most notably geographic area.

@@

The outputs of this model compared to the raw positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
Both show similar trends
The MRP estimate tends to increase the estimated positivity rate, compared to uncorrected positivity rate, because underrepresented groups tend to have higher positivity.

@@

Interpreting positivity, or prevalence, directly is difficult because it is a delayed and smoothed measure of incidence (see \cref{inc-prev:sec:prevalence-process}).
The major aim of this thesis is the ability to produce estimates of incidence, which can be interpreted in terms of population behaviour.

@@

Positive test results need to be grouped into \emph{infection episodes}.
An infection episode is a series of positive tests that originate from a single transmission event.
I use a method developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
Denote the time of a positive test by $t$, the number of consecutive negative tests prior to $t$ by $n$, and the time of the first positive test in the same individual by $t_p$ ($t_p = -\infty$ if there is no previous positive test in the same individual).
The test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
\begin{enumerate}
  \item $t_p = -\infty$.
  \item $n \geq 1$ and $t - t_p > 120$.
  \item $n \geq 2$ and $t - t_p > 90$.
  \item $n \geq 3$ and $t - t_p > 60$.
  \item $n \geq 4$.
  \item A complex heuristic based on the likely variant of the positive tests (see \textcite{weiRisk} for details). This heuristic rarely applied in the pre-Omicron era~\citePersonalComms{Sarah Walker}.
\end{enumerate}

A negative between two positive tests, when those positives are classed as part of the same infection episode, is known as an \emph{intermittent negative}.
Intermittent negatives are the clearest example of a false negative.
An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter, but demonstrate that false negatives do occur.

A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results.
As can be seen, the number of tests within an episode, their length, and other characteristics vary greatly.
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-5cm}
  \makebox[\textwidth][c]{\includegraphics[height=.9\paperheight]{biology-data/STATS13734/data_viz}}
  \caption[CIS infection episodes]{%
    500 randomly selected infection episodes.
    Shown are the negative tests before and after the outermost positive tests associated with the episode.
    Red dots are negative tests, green dots are positive tests, and blue dots are void tests (where the result could not be determined; excluded from other analyses).

    This figure has been provided as management information for operational planning purposes. They are provisional estimates circulated before public release. This management information should not be shared widely and should only be used for operational planning purposes. This information is not to be used publicly and ahead of being released publicly by ONS.
  }
\end{figure}
  \todo{Work out what to do with this figure....}

>>>>

@@

In this thesis, I use data from two studies which recruited participants from the general population in the UK.
Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
Therefore, the results from these studies are more generalisable to the entire UK population.

\todo[inline]{Some of this might be moved into the introductory chapter motivating each part of the thesis.}

The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.

The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest.
However, the results are limited because the sparse sampling which means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.

@@

The \emph{positivity rate} is the proportion of people in a given population that (if they were tested) would test positive for SARS-CoV-2 at a defined point in time; in this context test means an RT-PCR test~\autocite{cisMethodsONS}.
A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
Larger time periods (\ie weeks instead of days) allow more granular estimates in other dimensions, most commonly geography.

@@

The outputs of this model compared to the raw positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
Both show similar trends.
The MRP estimate most often increases the estimated positivity rate, compared to uncorrected positivity rate, because the underrepresented demographic groups tended to have higher positivity rates.

@@

Interpreting positivity, or prevalence, directly is difficult because it is a delayed and smoothed measure of incidence (see \cref{E-inc-prev:sec:prevalence-process}).
The major aim of this thesis is the ability to produce estimates of incidence, which can be interpreted in terms of population behaviour.

@@

Positive test results need to be grouped into \emph{infection episodes}.
An infection episode is a series of positive tests that originate from a single transmission event.
I use a heuristic-based algorithm developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.

The algorithm determines whether a new positive tests, at time $t$, is part of the previous infection episode in the same individual, where that previous infection episode had its first positive test at $t_p$.
If the individual has no previous positive tests, the new positive test is always the start of a new infection episode.
Otherwise, the test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
\begin{enumerate}
  \item $n \geq 1$ and $t - t_p > 120$.
  \item $n \geq 2$ and $t - t_p > 90$.
  \item $n \geq 3$ and $t - t_p > 60$.
  \item $n \geq 4$.
  \item A complex heuristic based on the likely variant of the positive tests (see \textcite{weiRisk} for details). This heuristic rarely applied in the pre-Omicron era~\citePersonalComms{Sarah Walker}.
\end{enumerate}

A negative between two positive tests, when those positives are classed as part of the same infection episode, is known as an \emph{intermittent negative}.
Intermittent negatives are the clearest example of a false negative.
I will generally consider that intermittent negatives are uninformative on the duration of detectability, and hence exclude them from several analyses.

A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results shown in \cref{biology-data:fig:episodes}.
As can be seen, the number of tests within an episode, their length, and other characteristics vary greatly.
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-5cm}
  \makebox[\textwidth][c]{\includegraphics[height=.9\paperheight]{biology-data/STATS13734/data_viz}}
  \caption[CIS infection episodes]{%
    500 randomly selected infection episodes.
    Shown are the negative tests before and after the outermost positive tests associated with the episode.
    Red dots are negative tests, green dots are positive tests, and blue dots are void tests (where the result could not be determined; excluded from other analyses).

    This figure has been provided as management information for operational planning purposes. They are provisional estimates circulated before public release. This management information should not be shared widely and should only be used for operational planning purposes. This information is not to be used publicly and ahead of being released publicly by ONS.
  }
  \label{biology-data:fig:episodes}
\end{figure}
  \todo{Work out what to do with this figure....}
